Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-22 Paper/paper.tex
removed fig.10 and discussion -- uninteresting

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\D}{\mathrm{d}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\dlambda}{\delta\Lambda}

@@

strength (SNR) required to constrain $\lambdans$ of the companion NS within a factor
of $100\%$ of its true value, i.e. with $\pm 50\%$ error-bars. The NS mass is fixed
at $1.35M_\odot$, spin at zero, and $\lambdans=2000$.

@@

We can see that, irrespective of BH properties, we cannot constrain $\lambdans$
to better than $\pm 50\%$ if the SNR is below $\approx 29$.

@@

As an illustration, we show the recovered
probability distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix $q = m_1/m_2 =
5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while varying $\lambdans$
over $\{1000, 1500, 2000\}$ between the three panels. The SNR is fixed at $\rho=50$.
The true value of $\lambdans$ is given the green (dashed) vertical line,
while the median recovered value is marked by a red (dashed) line. The darker
shaded region marks the $90\%$ confidence interval on $\lambdans$. One thing
to note in this figure is that the measurement uncertainty on $\lambdans$, i.e.
the full width of its $90\%$ confidence interval, grows slower than $\lambdans$
as $\lambdans$ is turned up (while the fractional uncertainty 
$(\Delta\lambdans)^{90\%}/\lambdans$ decreases with increasing $\lambdans$).
%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
$\rho=30$ (instead of $50$). In the top panel, we show the correlation between
non-tidal parameters, as manifested in their recovered joint probability
distributions.
In the middle row, we show the same between $\lambdans$ and non-tidal 
parameters, i.e. chirp mass (left panel), mass ratio $\eta$ (middle panel), and
BH spin $\chibh$ (right). The same is shown in the row below, with the
difference that the SNR has been turned up to $50$ to bring out the intrinsic
ambiguity contours.
%
We find that the measurement of $\lambdans$ is only weakly degenerate with
other parameters, and may not improve much if we knew non-tidal parameters
to better accuracy. The predominant factor that would improve the measurement
accuracy for $\lambdans$ is the signal strength, SNR.

@@

The precision with which $\lambdans$ can be measured is given by the width of 
its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

In Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we take the case of
moderately deformable neutron stars, with actual $\lambdans = 1000$ (left 
panel) and $1500$ (right panel), and ask how loud such a signal will have
to be for us to put meaningful constraints on the tidal deformability of NSs.
In the left panel, as a function BH mass and spin, 
we show the lowest SNR value $\rho_c$ in order for the signal to have enough
information that we can constrain $\dlambda$ below $200\%$.
\red{%
\begin{enumerate}
 \item Inconsistent notation $\delta$ vs $\dlambda$. Unify.
 \item Explain precise definition of $\delta\lambdans/\dlambda$: Is this the 
 full width or half width of the posterior. Such explanation fit in nicely into
 the discussion of the new $\lambdans$ posterior plots like Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}.
\end{enumerate}
}
\hspace{-3mm}Note that being able to constrain $\lambdans$ within a factor of $2$ interval 
(i.e. within $\pm 100\%$) is a fairly modest requirement. We find promising 
results, which suggest that there is a fairly wide range of BH spins, less
wide (but non-trivial) range of BH masses, within
which if we get a moderately loud NSBH signal with $\rho\approx 23$, we constrain
$\lambdans$ within a factor of $2$ interval (i.e. within $\pm 100\%$) around
the true value. While this is not very precise, it can possibly allow us to
rule out candidates for the equation of state of neutron star matter.
% 
If we want to constrain $\lambdans$ better with a single observation, however,
we need SNRs of $30$ or higher. This can be seen from the right
panel of Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}, which is similar
to the left but for a higher $\lambdans (=1500)$ system, with a more stringent
constraint requirement that $\lambdans$ be measured within a factor of $1$ of 
its true value (i.e. with $\pm 50\%$ errors, instead of $\pm 100\%$).
\red{Dont use two different thresholds in figure.} Even for
the most conducive cases, we {\it need} the SNR to be above $30$ to reach our 
accuracy threshold (which is more than $3\times$ less likely to happen than
getting a signal with $\rho=20$).

\red{%
(i) Rework text, so that you slowly build up the information needed. First, explain
the general idea pursued, then discuss one panel first, then introduce second
panel, finally, contrast both panels.
(ii) High information density. Distinction between left and right panels notation
not needed in the first sentence. deferring this info by a few sentences will make
it easier for the reader to follow the discussion.
}
\prayush{%
Lastly, in Fig.~\ref{fig:TT_LambdaErrorCurves_BHspin_BHmass_CI90_0} we show the
conditions that BH parameters must fulfill beside having an SNR of $20$ (left 
panel) or $30$ (right panel), to obtain meaningful constraints on $\lambdans$.
At $\rho=20$, the left panel shows that we need high BH spins 
$\chibh\gtrsim +0.6$ and a very deformable NS with $\lambdans\geq 1500$, in 
order to put a factor of $2$ bound on $\lambdans$. At $\rho=30$, with
$\lambdans\gtrsim 1000$ and $\chibh\geq+0.6$ we can put a similar factor of
$2$ bound on $\lambdans$, but with $\chibh\sim+0.75$ and an extremely deformable
NS with $\lambdans=2000$ one could even put a $\pm 50\%$ error-bar on 
$\lambdans$ measurement.
}

>>>>

% Macros for text notes and comments
\newcommand{\Note}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\D}{\mathrm{d}}
\newcommand{\lambdans}{\Lambda_\mathrm{NS}}
\newcommand{\dlambda}{\delta\lambdans^{90\%}}

@@

strength (SNR) required to constrain $\lambdans$ within an interval of width
equal to $100\%$ of its true value, i.e. with $\pm 50\%$ error-bars. The NS mass
is fixed at $1.35M_\odot$, spin at zero, and $\lambdans=2000$.

@@

We can see that, even in the most conducive circumstances with large aligned 
$\chibh$ and a comparable mass BH, we can only constrain $\lambdans$ to better
than $\pm 50\%$ {\it if} the SNR is $\gtrsim 29$. In the era of design
sensitivity LIGO instruments, we expect this to happen approximately once in a
year of observation.

@@

As an illustration, we show the recovered
probability distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix $q = m_1/m_2 =
5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while varying $\lambdans$
over $\{1000, 1500, 2000\}$ between the three panels. The SNR is fixed at $\rho=50$.
The true value of $\lambdans$ is given the green (dashed) vertical line,
while the median recovered value is marked by a red (dashed) line. The darker
shaded region marks the $90\%$ confidence interval on $\lambdans$. One thing
to note in this figure is that the measurement uncertainty on $\lambdans$, i.e.
the full width of its $90\%$ confidence interval, grows slower than $\lambdans$
as $\lambdans$ is turned up (while the fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases with increasing $\lambdans$).
%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
$\rho=30$ (instead of $50$). In the top panel, we show the correlation between
non-tidal parameters, as manifested in their recovered joint probability
distributions.
In the middle row, we show the same between $\lambdans$ and non-tidal 
parameters, i.e. chirp mass (left panel), mass ratio $\eta$ (middle panel), and
BH spin $\chibh$ (right). The same is shown in the row below, with the
difference that the SNR has been turned up to $50$ to bring out the intrinsic
ambiguity contours.
%
We find that the measurement of $\lambdans$ is only weakly degenerate with
other parameters, and may only marginally improve if we knew non-tidal parameters
to better accuracy. The predominant factor that would improve the measurement
accuracy for $\lambdans$ is the signal strength, SNR.

@@

%The precision with which $\lambdans$ can be measured is given by the full 
%width of its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

In Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we take the case of
moderately deformable neutron stars, with actual $\lambdans = 1000$ (left 
panel) and $1500$ (right panel), and ask how loud such a signal will have
to be for us to put meaningful constraints on the tidal deformability of NSs.
In the left panel, as a function BH mass and spin, 
we show the lowest SNR value $\rho_c$ in order for the signal to have enough
information that we can constrain $\dlambda$ below $200\%$.
%\red{%
%\begin{enumerate}
% \item Inconsistent notation $\delta$ vs $\dlambda$. Unify.
% \item Explain precise definition of $\delta\lambdans/\dlambda$: Is this the 
% full width or half width of the posterior. Such explanation fit in nicely into
% the discussion of the new $\lambdans$ posterior plots like Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}.
%\end{enumerate}
%}
%\hspace{-3mm}
Note that being able to constrain $\lambdans$ within a factor of $2$ interval 
(i.e. within $\pm 100\%$) is a fairly modest requirement. We find promising 
results, which suggest that there is a fairly wide range of BH spins, less
wide (but non-trivial) range of BH masses, within
which if we get a moderately loud NSBH signal with $\rho\approx 23$, we constrain
$\lambdans$ within a factor of $2$ interval (i.e. within $\pm 100\%$) around
the true value. While this is not very precise, it can possibly allow us to
rule out candidates for the equation of state of neutron star matter.
% 
If we want to constrain $\lambdans$ better with a single observation, however,
we need SNRs of $30$ or higher. This can be seen from the right
panel of Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}, which is similar
to the left but for a higher $\lambdans (=1500)$ system, with a more stringent
constraint requirement that $\lambdans$ be measured within a factor of $1$ of 
its true value (i.e. with $\pm 50\%$ errors, instead of $\pm 100\%$).
\red{Dont use two different thresholds in figure.} Even for
the most conducive cases, we {\it need} the SNR to be above $30$ to reach our 
accuracy threshold (which is more than $3\times$ less likely to happen than
getting a signal with $\rho=20$).

%\red{%
%(i) Rework text, so that you slowly build up the information needed. First, explain
%the general idea pursued, then discuss one panel first, then introduce second
%panel, finally, contrast both panels.
%(ii) High information density. Distinction between left and right panels notation
%not needed in the first sentence. deferring this info by a few sentences will make
%it easier for the reader to follow the discussion.
%}
%\prayush{%
%Lastly, in Fig.~\ref{fig:TT_LambdaErrorCurves_BHspin_BHmass_CI90_0} we show the
%conditions that BH parameters must fulfill beside having an SNR of $20$ (left 
%panel) or $30$ (right panel), to obtain meaningful constraints on $\lambdans$.
%At $\rho=20$, the left panel shows that we need high BH spins 
%$\chibh\gtrsim +0.6$ and a very deformable NS with $\lambdans\geq 1500$, in 
%order to put a factor of $2$ bound on $\lambdans$. At $\rho=30$, with
%$\lambdans\gtrsim 1000$ and $\chibh\geq+0.6$ we can put a similar factor of
%$2$ bound on $\lambdans$, but with $\chibh\sim+0.75$ and an extremely deformable
%NS with $\lambdans=2000$ one could even put a $\pm 50\%$ error-bar on 
%$\lambdans$ measurement.
%}
